gap' issue

### 2017-06-13 analysis/ObjGap.rmd
Files commited by wflow_commit.

## introduction of the problem.

@@

### old version documnt

@@

In the document, we have several observations:

@@

## examples

@@

```{r}
sim_K = function(K, N, P, SF, SL, signal,noise){
  E = matrix(rnorm(N*P,0,noise),nrow=N)
  Y = E
  L_true = array(0, dim = c(N,K))
  F_true = array(0, dim = c(P,K))

@@

>>>>

## 0 introduction of the problem.

@@

### 0.1 old version documnt

@@

In the document, we have several observations (**please ignore this section since it is just a record for the old code, I will show all the cases in the next section**):

@@

## 1 summary of the result

- section 2.1 shows that in the greedy algorithm doesn't always increase the lower bound of the log-likelihood when we add one more factor. When we add one more factor the marginal (conditional) likelihood, $E_q \log\{P(Y|L,F)\} = -\frac{1}{2\sigma^2}(Y_{ij} - \sum_k l_{ik}f_{kj})^2$, would increase, but the $\log(\frac{P(L)}{q(L)}) = \sum_k \log(\frac{P(L_k)}{q(L_k)})$ would decrease since:
\begin{eqnarray} E_q log(\frac{P}{q}) \leq log E_q (\frac{P}{q}) = 0 \end{eqnarray}

> I think the KL(q || p) is not continuous as q,p -> delta_0 [point mass at 0]
> eg q_n = N(0,1/n) and p_n = N(0,2/n)
> I think lim_{n \rightarrow \inf} KL(q_n, p_n) may not be 0?
> If that is right then the lower bound F(g; qhat(g)) is not necessarily continuous in g (even though likelihood is continuous in g I think)…
> maybe this explains the gap; and may also tell us what to do about it.
> —— Matthew's comments

$$lim_{n \rightarrow \inf} KL(q_n, p_n) = -\frac{1}{2}\{\log2 - \frac{1}{2}\} \neq 0$$

- section 2.2 shows that rank one FLASH with 4 different method: `method = "fdr"`, `method = "shrink"`, `method = "spike and slab"`, `method = "uniform with point mass"`. We can see that the lower bound of the log likelihood is increase if we apply different methods without penalty term. The objective value is not always increase in `method = "fdr"` case.

- In section 2.2, there is an interesting observation that the RMSE results don't always agree wit the value of the objective function, which means that the method with larger objective function value doesn't has smaller RMSE value (better in prediction).

- In section 2.3 we use `method = "shrink"` to the same data in section 2.1.2 where the loglikelihood is not monotonically increase, and we find that the loglikelihood still decrease when we add fifth factor. So this is probably because of the penalty term.

- Section 2.4 shows that the `method = "fdr"` performs very well in rank zero case.

- In section 2.3 and 2.12, we can see that rank 4 model have larger value of lower bound but rank 5 model has better performance in RMSE.

## 2 examples

### 2.1 greedy algorithm

@@

```{r}
library(flashr)
sim_K = function(K, N, P, SF, SL, signal,noise){
  E = matrix(rnorm(N*P,0,noise),nrow=N)
  Y = E
  L_true = array(0, dim = c(N,K))
  F_true = array(0, dim = c(P,K))

@@

#### 2.1.1

"fdr" method looks fine in many cases.

```{r}
Greedy_showcase = function(K=5,N = 100, P = 200, SF = 0.8, SL = 0.8, signal = 1,noise = 1){
  set.seed(99)
  data = sim_K(K,N, P, SF , SL , signal ,noise )
  Y = data$Y
  E = data$Error
  g1 = greedy(Y,K = 10)
  plot(g1$clik_vec + cumsum(g1$priorpost_vec),main = "fdr method")
  # show the prediction
  rmse_5 = sqrt(mean((Y - g1$l %*% t(g1$f) -E)^2)) / sqrt(mean((Y - E)^2))
  rmse_4 = sqrt(mean((Y - g1$l[,1:4] %*% t(g1$f[,1:4]) -E)^2)) / sqrt(mean((Y - E)^2))
  return(list(obj_track = g1$clik_vec + cumsum(g1$priorpost_vec),RMSE = c(rmse_4,rmse_5)))
}
Greedy_showcase(K=5,N = 100, P = 200, SF = 0.8, SL = 0.8, signal = 1,noise = 1)
```

#### 2.1.2

But in some case, the lower bound is decrease when we add one more factor:

```{r}
Greedy_showcase(K=5,N = 100, P=200, SF = 0.5, SL = 0.5, signal = 1,noise = 3)
```

We guess that it is because the penalty term in `ash` we use.

### 2.2 rank one model with different method

Here we use four different methods: fdr, shrink, uniform without penalty and spike and slab.

```{r}
rank1test = function(K,N, P, SF , SL , signal ,noise ){
  set.seed(99)
  par(mfrow = c(3, 2))
  par(cex = 0.6)
  par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
  data = sim_K(K,N, P, SF , SL , signal ,noise )
  Y = data$Y
  E = data$Error
  # "l" is for lower bound
  g3_1 = flash(Y,objtype = "l")
  plot(g3_1$obj_val_track,main = "fdr method",type = "l")
  g3_2 = flash(Y,objtype = "l",ash_para = list( prior = "uniform",pointmass = TRUE,method = NULL))
  plot(g3_2$obj_val_track,main = "uniform + pointmass method",type = "l")
  g3_3 = flash(Y,objtype = "l",ash_para = list(method = "shrink"))
  plot(g3_3$obj_val_track,main = "shrink method",type = "l")
  g3_4 = flash(Y,objtype = "l",ash_para = list( prior = "uniform",pointmass = TRUE,method = NULL,mixsd = c(1)))
  plot(g3_4$obj_val_track,main = "spike and slab method",type = "l")
  obj = c(g3_1$obj_val,g3_2$obj_val,g3_3$obj_val,g3_4$obj_val)
  RMSE = c(sqrt(mean((Y - g3_1$l %*% t(g3_1$f) -E)^2)) / sqrt(mean((Y - E)^2)),
           sqrt(mean((Y - g3_2$l %*% t(g3_2$f) -E)^2)) / sqrt(mean((Y - E)^2)),
           sqrt(mean((Y - g3_3$l %*% t(g3_3$f) -E)^2)) / sqrt(mean((Y - E)^2)),
           sqrt(mean((Y - g3_4$l %*% t(g3_4$f) -E)^2)) / sqrt(mean((Y - E)^2)))
  plot(RMSE, col =c("magenta","cyan","yellow","green"), pch = 16, cex = 2,main = "RMSE")
  legend("top", cex = 1,
       legend = c("fdr","Uniform","shrink","SpikeSlab"),
       col = c("magenta","cyan","yellow","green"),
       pch=16)
  plot(obj, col =c("magenta","cyan","yellow","green"), pch = 16, cex = 2,main = "objective value")
  legend("top",  cex = 1,
       legend = c("fdr","Uniform","shrink","SpikeSlab"),
       col = c("magenta","cyan","yellow","green"),
       pch=16)
}

rank1test(1,100,200,0.5,0.5,1,1)
```

we can see that the RMSE doesn't agree with the value of objective function.

```{r}
rank1test(1,100,200,0.8,0.8,1,2)
```

We can see that the objective function is not increasing in `method = "fdr"`

```{r}
rank1test(1,20,30,0.5,0.6,1,1)
```

#### very sparse case

```{r}
rank1test(1,100,300,0.9,0.9,1,1)
```

```{r}
rank1test(1,100,300,0.95,0.99,1,0.5)
```

### 2.3 apply greedy without penalty

```{r}
Greedy_showcase_1 = function(K=5,N = 100, P = 200, SF = 0.8, SL = 0.8, signal = 1,noise = 1,flash_para = list(ash_para = list(method = "shrink"))){
  set.seed(99)
  data = sim_K(K,N, P, SF , SL , signal ,noise )
  Y = data$Y
  E = data$Error
  g1 = greedy(Y,K = 10,flash_para)
  plot(g1$clik_vec + cumsum(g1$priorpost_vec),main = "shrink method")
  # show the prediction
  rmse_5 = sqrt(mean((Y - g1$l %*% t(g1$f) -E)^2)) / sqrt(mean((Y - E)^2))
  rmse_4 = sqrt(mean((Y - g1$l[,1:4] %*% t(g1$f[,1:4]) -E)^2)) / sqrt(mean((Y - E)^2))
  return(list(obj_track = g1$clik_vec + cumsum(g1$priorpost_vec),RMSE = c(rmse_4,rmse_5)))
}
Greedy_showcase_1(K=5,N = 100, P=200, SF = 0.5, SL = 0.5, signal = 1,noise = 3,flash_para = list(ash_para = list(method = "shrink")))
```

by using the `method = "shrink"`, we can see that the objective function getting large but the lower bound is still not monotonically increasing. So it's not because of the penalty we use in `ash`.

### 2.4 rank zero case

```{r}
rank1test_0 = function(K,N, P, SF , SL , signal ,noise ){
  set.seed(99)
  par(mfrow = c(3, 2))
  par(cex = 0.6)
  par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
  data = sim_K(K,N, P, SF , SL , signal ,noise )
  Y = data$Y
  E = data$Error
  # "l" is for lower bound
  g3_1 = flash(Y,objtype = "l")
  plot(g3_1$obj_val_track,main = "fdr method",type = "l")
  g3_2 = flash(Y,objtype = "l",ash_para = list( prior = "uniform",pointmass = TRUE,method = NULL))
  plot(g3_2$obj_val_track,main = "uniform + pointmass method",type = "l")
  g3_3 = flash(Y,objtype = "l",ash_para = list(method = "shrink"))
  plot(g3_3$obj_val_track,main = "shrink method",type = "l")
  g3_4 = flash(Y,objtype = "l",ash_para = list( prior = "uniform",pointmass = TRUE,method = NULL,mixsd = c(1)))
  plot(g3_4$obj_val_track,main = "spike and slab method",type = "l")
  obj = c(g3_1$obj_val,g3_2$obj_val,g3_3$obj_val,g3_4$obj_val)
  RMSE = c(sqrt(mean((Y - g3_1$l %*% t(g3_1$f) -E)^2)) ,
           sqrt(mean((Y - g3_2$l %*% t(g3_2$f) -E)^2)) ,
           sqrt(mean((Y - g3_3$l %*% t(g3_3$f) -E)^2)) ,
           sqrt(mean((Y - g3_4$l %*% t(g3_4$f) -E)^2)) )
  plot(RMSE, col =c("magenta","cyan","yellow","green"), pch = 16, cex = 2,main = "RMSE")
  legend("top", cex = 1,
       legend = c("fdr","Uniform","shrink","SpikeSlab"),
       col = c("magenta","cyan","yellow","green"),
       pch=16)
  plot(obj, col =c("magenta","cyan","yellow","green"), pch = 16, cex = 2,main = "objective value")
  legend("top",  cex = 1,
       legend = c("fdr","Uniform","shrink","SpikeSlab"),
       col = c("magenta","cyan","yellow","green"),
       pch=16)
}

rank1test_0(1,20,60,1,0.6,1,1)
```

```{r}
rank1test_0(1,100,200,1,0.6,1,2)
```

```{r}
rank1test_0(1,60,100,1,0.6,1,0.5)
```
